An anthropological account of the Vim editor:\\ ouroboros, features and tweaks after 10 years of usage

### 2017-11-09 article/article.tex
finished first version of article

\subsubsection{Ex}
This might be the least popular mode.
One might use \tttt{q:,/,?} to have a window
with colon or search command history to be edited normaly,
and the chosen command can be run with \ttt{CR}.
Vimscript was largerly based on the ex editor language~\cite{ex},
and a more advanced user might use it for prototyping
by defining mappings and settings and managing sessions and scripts
propably in a \tttt{plugin/} folder of a directory in \tttt{:echo \&runtimepath}.
In the default interface started with the \tttt{Q} command in normal mode,
each command is input without entering : again. Use \tttt{:vi} to exit Ex mode,
follow documentation from \tttt{:h Ex-mode} for further information.

@@

\subsection{Color}\label{visual2}
Newer versions of Vim support 24 bit true color (aka 16 million colors)
in terminal Vim (not only in GVim).
The terminal must support true color, and tests are available e.g. in
\url{https://gist.github.com/XVilka/8346728}.
Then Vim needs to be set to use true colors with
\tttt{:set termguicolors}.
If using 8 or 16 bit colors, Vim uses the color pallete from the
terminal, if using true colors each color is defined directly.
Settings for using true colors inside Byobu/Tmux involve tweakings
and are available in~\cite{vimrc}.
Good color schemes to use with true colors are Gruvbox~\cite{gruv}
and Solarize~\cite{sol}.
One might source syntax files at any time to change syntax, usually
though linking tokens to syntax groups with
\tttt{:syntax keyword <group\_name> <token1> <token2> <token3> ...}
and then relating the group to another group 
\tttt{:highlight link <group\_name> <group\_name2>}
or by specifying the group characteristics directly:
\tttt{highlight <group\_name> guifg=\#ffffff}.
If you change a syntax file, reloading a file with with \tttt{:e<CR>}
updates the highlighting on the window with the corresponding filetype.
A complete syntax highlighting support typically involves at least three files:
\begin{itemize}
  \item \tttt{ftdetect/<filetype>.vim}, where the file type is
    detected with e.g.
    \tttt{:autocmd BufNewFile,BufRead *.<file\_extension> setfiletype <filetype>}
  \item \tttt{ftplugin/<filetype>.vim} with general settings for the
    file type, such as: 
    \tttt{:set tabstop=2 softtabstop=2 shiftwidth=2 textwidth=70
    expandtab autoindent}.
  \item A \tttt{syntax/<filetype>.vim} file, with bindings between
    tokens and highlighting groups; and highlighting group definitions.
\end{itemize}

@@

\noindent {\bf Potential next steps} in using Vim:
\begin{itemize}
  \item Measure the performance of text mining routines in Vim against those implemented in C or Python.
  \item Enhance the HTTP browsing capabilities of~\cite{vimrc}.
  \item Better integrate Python and Vimscript, especially for data visualization
    and syntax highlighting management.
    In accordance with the visualization issues described in
    Sections~\ref{visual1} and~\ref{visual2}.
  \item Make plugins for:
    \begin{itemize}
      \item listing all the mappings available in each mode and the typing combinations which are available
        for new mappings.
      \item Sessions, as described in Section~\ref{visual1}.
      \item AA messages (shouts): to keep track, document and share of working sessions
        as in~\cite{aa1,aa2}, with capabilities to manage AA sessions,
        send visual or sonic cues for temporal marks, use Vim state to build AA shouts,
        relate AA sessions to other media, such as software repositories,
        screencasts and images, interact with IRC channels and other social platforms.
      \item Slick Vim: a collection of the settings, mappings and plugins I use.
        Enhancements such as using \ttt{C-} commands also to browse tabs,
        shortcuts to join a window into a tab,
        and dummy minimal plugins as simplest, file type and syntax highlighting,
        Some more elaborate tweaks should also be present, such
        as breaking lines in sensible places for natural language texts
        while respecting e.g. \tttt{:se textwidth}.
      \item Dealing with .swo and .swp temporary files.
        In summary, if the restored .swo file has the same content
        and the file being opened,
        the restoration phase can be omitted.
        If the contents differ, Vim should open a tab with each file
        in a vertical split and run \tttt{:windo :diffthis}.
      \item Rendering images and equations.
        These are useful for using Vim in presentations
        or achieving a textual representation when it is mandatory,
        such as to comply with the limitations of a platform (e.g. Vim editor). 
        but also hold stylistic merits as ascii art is often
        very appreciated.
        One can both obtain an ascii representation of a binary image (e.g. JPG, PNG),
        and can directly render ascii charts from data using cues such are shape, position
        and color..
      \item Slide presentations. I've been using some automation for browsing
        slides and opening figures and some Vim users asked for the settings and
        commands. They are very elementary use of registers being executed over
        arbitrary but consistent textual patterns.
      \item Changing the color scheme an highlighting scheme incrementally
        and selectively using the features described in Section~\ref{visual2}.
    \end{itemize}
\end{itemize}

\subsection*{Acknowledgments}
FAPESP, Vim developers and documentation maintainers,
Vim user community.

\appendix
\section{Example of usage session}
% open file, split, tab, find, replace char,
% append to line, visual block, etc.
\section{Key naming in Vim documentation}
% CTRL, <C-X> <C-S-X>s 1250
% Special chars..
\section{My vimrc file and usage}
In this Appendix is my vimrc file.
Although it has comments, one should
look for the options that (s)he does
not understand, as a thorough explanation
of the file is tedious and out of the scope
of this document.

Mostly, I toggle the status line with \texttt{\\\\B}
and the tab line with \texttt{\\\\T}.
Save and close windows with \texttt{\\w} and \texttt{\\q}.

\section{Example of notations on mappings}\label{notes}
\texttt{:h index} shows all the mappings.
\subsection{Normal mode}
Every letter and character in the keyboard is used.
In Normal mode: <TAB> is the same as <C-I>,
<BS> and <C-H> is the same as h (<C-H> don't sowk here).
<C-J> and <C-N> is the same as j.
C-P is the same as k.
Space is same as l
CTRL-[ <Esc> are not used
C-\ a-z reserved for extensions
CTRL-\_ not used

learn how to use filters ! and !!
% .%#: registers
\url{https://github.com/vim/vim/blob/master/runtime/doc/index.txt}.

my usage of <C-(HJKL)> for jumping through windows.
<C-L> is lost though, and require :redraw or :e

+ same as <CR> which is not that useful.

Many ][ combinations are not used, e.g.
with abhjklfg
\_ might be used as a more powerful \^, leaving it free for mappings.

@@

Del is same as x

text objects: redundancy between a/i(),{},[],<>

many entries available through the g commands
z for folds, spell checking and some movements (mainly when wrap is set).

\subsection{Insert mode}
All standard char keys are used for entering chars on text.
<C-G>jk can be achieved by <C-O>jk which is more powerful
in moving through multiple lines.
j-[ same as ESC
		CTRL-\ a - z	reserved for extensions
		CTRL-\ others	not used

<C-J> and <C-M> are <CR>

\section{Overview of my usage of Vim}
% move to usage appendix TTM
in some of the basic topics, such as in using numerous terminals 
(it works pretty well!) and using sessions (Section~\ref{sessions}).
These make Vim very convenient e.g. for users that edit multiple files,
run scripts and compile programs and latex files.
I used for years multiple terminal abs and windows,
and sometimes a byobu/screen terminal multiplexer.
Nowadays I am using standard Vim windows as terminals
such as described in Section~\ref{terminals}.

\section{My \tttt{:version}}

% sessions, terminal for ipython3 and evince
% alternative with byobu or many terminals
% historical usage, other editors,
% most often edited file types

\end{document}

>>>>

\subsubsection{Ex}
This might be the least popular mode.
One might use \tttt{q:,/,?} to have a window
with colon or search command history to be edited normaly,
and the chosen command can be run with \ttt{CR}.
Vimscript was largerly based on the ex editor~\cite{ex},
and a more advanced user might use it for prototyping
by defining mappings and settings and managing sessions and scripts
propably in a \tttt{plugin/} folder of a directory in \tttt{:echo \&runtimepath}.
In the default interface started with the \tttt{Q} command in normal mode,
each command is input without entering : again. Use \tttt{:vi} to exit Ex mode,
follow documentation from \tttt{:h Ex-mode} for further information.

@@

\subsection{Color}\label{visual2}
Newer versions of Vim support 24 bit true color (aka 16 million colors)
in terminal Vim (not only in GVim).
The terminal must support true color, and tests are available e.g. in
\url{https://gist.github.com/XVilka/8346728}.
Then Vim needs to be set to use true colors with
\tttt{:set termguicolors}.
If using 8 or 16 bit colors, Vim uses the color pallete from the
terminal, if using true colors each color is defined directly.
Settings for using true colors inside Byobu/Tmux involve tweakings
and are available in~\cite{vimrc}.
Good color schemes to use with true colors are Gruvbox
and Solarize.
One might source syntax files at any time to change syntax, usually
though linking tokens to syntax groups with
\tttt{:syntax keyword <group\_name> <token1> <token2> <token3> ...}
and then relating the group to another group 
\tttt{:highlight link <group\_name> <group\_name2>}
or by specifying the group characteristics directly:
\tttt{highlight <group\_name> guifg=\#ffffff}.
If you change a syntax file, reloading a file with with \tttt{:e<CR>}
updates the highlighting on the window with the corresponding filetype.
A complete syntax highlighting support typically involves at least three files:
\begin{itemize}
  \item \tttt{ftdetect/<filetype>.vim}, where the file type is
    detected with e.g.
    \tttt{:autocmd BufNewFile,BufRead *.<file\_extension> setfiletype <filetype>}
  \item \tttt{ftplugin/<filetype>.vim} with general settings for the
    file type, such as: 
    \tttt{:set tabstop=2 softtabstop=2 shiftwidth=2 textwidth=70
    expandtab autoindent}.
  \item A \tttt{syntax/<filetype>.vim} file, with bindings between
    tokens and highlighting groups; and highlighting group definitions.
\end{itemize}

@@

\noindent {\bf Potential next steps} in using Vim:
\begin{itemize}
  \item Measure the performance of text mining routines in Vim against those implemented in C or Python.
  \item Enhance the HTTP browsing capabilities of~\cite{vimrc}.
  \item Better integrate Python and Vimscript, especially for data visualization
    and syntax highlighting management.
    In accordance with the visualization issues described in
    Sections~\ref{visual1} and~\ref{visual2}.
  \item Make plugins for:
    \begin{itemize}
      \item listing all the mappings available in each mode and the typing combinations which are available
        for new mappings.
      \item Sessions, as described in Section~\ref{visual1}.
      \item AA messages (shouts): to keep track, document and share of working sessions
        as in~\cite{aa1,aa2}, with capabilities to manage AA sessions,
        send visual or sonic cues for temporal marks, use Vim state to build AA shouts,
        relate AA sessions to other media, such as software repositories,
        screencasts and images, interact with IRC channels and other social platforms.
      \item Slick Vim: a collection of the settings, mappings and plugins I use.
        Enhancements such as using \ttt{C-} commands also to browse tabs,
        shortcuts to join a window into a tab,
        and dummy minimal plugins as simplest, file type and syntax highlighting,
        Some more elaborate tweaks should also be present, such
        as breaking lines in sensible places for natural language texts
        while respecting e.g. \tttt{:se textwidth}.
      \item Dealing with .swo and .swp temporary files.
        In summary, if the restored .swo file has the same content
        and the file being opened,
        the restoration phase can be omitted.
        If the contents differ, Vim should open a tab with each file
        in a vertical split and run \tttt{:windo :diffthis}.
      \item Rendering images and equations.
        These are useful for using Vim in presentations
        or achieving a textual representation when it is mandatory,
        such as to comply with the limitations of a platform (e.g. Vim editor). 
        but also hold stylistic merits as ascii art is often
        very appreciated.
        One can both obtain an ascii representation of a binary image (e.g. JPG, PNG),
        and can directly render ascii charts from data using cues such are shape, position
        and color.
      \item Redirecting the commands that usually show the results in a 'more' interface,
        which cannot be searched nor copied,
        to a parsed and linked quickfix or location window.
        The basic idea is to use \tttt{:redir} command to redirect the output
        of such commands with \tttt{:se nomore}.
      \item Slide presentations. I've been using some automation for browsing
        slides and opening figures and some Vim users asked for the settings and
        commands. They are very elementary use of registers being executed over
        arbitrary but consistent textual patterns.
      \item Changing the color scheme an highlighting scheme incrementally
        and selectively using the features described in Section~\ref{visual2}.
    \end{itemize}
\end{itemize}

\subsection*{Acknowledgments}
FAPESP (project 2017/05838-3); Vim developers and documentation maintainers;
Vim user community.

\appendix
\section{Example of usage session}
I usually begin by opening a file or directory
with \tttt{\$ vim <filename>}.
The color scheme is alternated between
blue and GruvBox with
\tttt{:colo gruvbox} and \tttt{:colo blue}.
I open an vertical split and then move
the window to a new tab using
\tttt{:vs} and \ttt{C-W\_T}.
I then search tokens related to
the enhancements I want to make or
the knowledge I want to acquire.
I go back to the previous tab with \tttt{gr}
and make a global replace with
\tttt{:\%s/<this>/<that/g}.
On adding dots to sentences,
I record in the \tttt{"q} register
the sequence \tttt{jA.jj},
using it as a macro 10 times by
\tttt{10@q}.
I move to the other tab with \tttt{gt}
and open a terminal window with \tttt{:term}
for compiling latex files.
I start another terminal with \tttt{<C-W> :term}
for opening the resulting PDFs with evince.
If any new idea comes to mind and I have time,
\tttt{\textbackslash\textbackslash s} opens 
my vimrc~\cite{vimrc} for editing and \tttt{\textbackslash s}
sources it.
If there is e.g. code or notes in other projects I am working
on, I reach them through \tttt{<Alt-arrows>} as other Vim instances
in Byobu/Tmux sessions and windows.

\section{My vimrc file and usage}
In this Appendix is my vimrc file.
Although it has comments, one should
look for the options that (s)he does
not understand, as a thorough explanation
of the file is tedious and out of the scope
of this document.

In using Vim with my vimrc file~\cite{vimrc},
I mostly toggle the status line with \texttt{\textbackslash\textbackslash B}
and the tab line with \texttt{\textbackslash\textbackslash T}.
Save and close windows with \texttt{\textbackslash w} and \texttt{\textbackslash q}.
The mappings for transitioning through splits and tabs
are also used constantly.

\section{Example of notes on mappings}\label{notes}
\texttt{:h index} shows all the default mappings
while \tttt{:map} shows the user-defined mappings.
By considering such information, one can make
useful observations exemplified in this Appendix.

\subsection{Normal mode}
Every letter and character in the keyboard is used.
In Normal mode: \ttt{TAB} is the same as \ttt{C-I},
\ttt{BS} and \ttt{C-H} is the same as h.
\ttt{C-J} and \ttt{C-N} is the same as j.
\ttt{C-P} is the same as k.
Space is same as l.
\ttt{C-[} and \ttt{Esc} are not used
\tttt{<C-\textbackslash> a-z} is reserved for extensions
\ttt{C-\_} not used.
\tttt{+} is the same as \ttt{CR} and they are both not very useful.
Del is same as x.

Many \tttt{][} combinations are not used, e.g.
with \tttt{abhjklfg}.
The \tttt{\_} command might be used as a more powerful \tttt{\^},
leaving it free for mappings.

@@

The \ttt{C-(HJKL)} commands are redundant,
with the exception of \ttt{C-L} which redraws the screen,
so it is a reasonable choice to use them to move focus
of the editor to splits in the \tttt{hjkl} directions.

Many key combinations are available for new mappings through the \tttt{g}
and \tttt{z}, and \tttt{v} commands.
They have typical uses, e.g. \tttt{z} for folds,
spell checking and some movements (mainly when wrap is set).

\subsection{Insert mode}
All standard char keys are used for entering chars on text.
\tttt{<C-G>(j,k)} can be achieved by \tttt{<C-O>(j,k)} which is more powerful
in moving through multiple lines.
\ttt{C-[} is the same as \ttt{ESC}.
\ttt{C-J} and \ttt{C-M} are \ttt{CR}.
\tttt{<C-\textbackslash> a-z} is reserved for extensions,
other combinations with \ttt{C-\textbackslash} are not used.

\section{My \tttt{:version}}
I should keep the output of \tttt{:version} executed on the system
in which I wrote this document in this link:
\url{}.
It was compiled with this Makefile~\cite{makefile}
in a 16.04 LTS Ubuntu Linux,
and is tagged as 8.0, Included Patches 1-1173.

\begin{thebibliography}{99}
\fontsize{11}{0}\selectfont
\bibitem{anPh}
	Anthropological physics and social psychology in the critical research of networks. Complex Networks Digital Campus (CS-DC'15).
	Available at \url{https://youtu.be/oeOKYc3-nbM}
\bibitem{anPh2}
	Fabbri, R. What are you and I? [anthropological physics fundamentals], 2015. Available at \url{https://www.academia.edu/10356773/What_are_you_and_I_anthropological_physics_fundamentals_}
\bibitem{ouroWiki}
  Ouroboros. (2017, November 2). In Wikipedia, The Free Encyclopedia. Retrieved 22:19, November 9, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Ouroboros&oldid=808392809}
\bibitem{vimrc}
	Fabbri, R. (2017). A reasonable vimrc file. Available at \url{https://raw.githubusercontent.com/ttm/vim/master/vimrc} 
\bibitem{ex}
  Ex (text editor). (2017, March 22). In Wikipedia, The Free Encyclopedia. Retrieved 22:22, November 9, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Ex_(text_editor)&oldid=771621020}
\bibitem{tokipona}
	Fabbri, R. (2017). A Toki Pona Python Package and Vim Syntax Highlighting. Available at \url{https://github.com/ttm/tokipona} 
\bibitem{tpLang}
	Lang, S. (2014). Toki Pona: the language of good. Tawhid Publishing.
    ISBN-10: 0978292308, ISBN-13: 978-0978292300.
\bibitem{aa1}
	Fabbri, R., Fabbri, R., Vieira, V., Penalva, D., Shiga, D., Mendonça, M., Negrao, A., Zambianchi, L., \& Thumé, G. (2013). AA: The Algorithmic Autoregulation (Distributed Software Development) Methodology. RESI. From \url{https://arxiv.org/abs/1604.08255}
\bibitem{aa2}
	Fabbri, R. (2017).
The Algorithmic-Autoregulation (AA) Methodology and Software:
a collective focus on self-transparency. ENMC2017. From \url{https://github.com/ttm/ensaaio/raw/master/emc/article.pdf} 
\bibitem{makefile}
	Fabbri, R. (2017). The Makefile with which I compiled Vim for this article. Available at \url{https://raw.githubusercontent.com/ttm/vim/master/Makefile} 
\end{thebibliography}
\end{document}
